\\Large The heat is on: Rising temperatures alter how and where mammals move

### 2025-09-08 writing/manuscript.Rmd
reviewed discussion; need to finish "future work"

Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010]. While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more severe physiological damage in a shorter span of time [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024]. Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, especially with increasing body size [@dyer_travel_2023; @verzuh_behavioural_2023]. Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature genralists') can adapt more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024]. Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond within a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].

@@

In the last century, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ipcc_climate_2023]. As heat stress intensifies over the next 100 years, mammals will continue to suffer impacts on their fitness, movement rates, and habitat selection [@deb_modelling_2020; @woo-durand_increasing_2020], which will be exacerbated by more frequent and intense extreme events [@bunnell_global_2011] and anthropogenic habitat loss, modification, and fragmentation [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024]. It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018]. Preparing for and responding to future changes require a comprehensive understanding of how temperature will affect how and where mammals move, as well as the implications such changes will have on species' adaptability and overall community structure. Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behaviour [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [e.g., body size, activity time, and elevational and latitudinal distribution @mccain_assessing_2019]. However, there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behaviour and quantitative predictions of how their behaviour will respond to climate change in the current century [@ipcc_climate_2023].

@@

Western Canada is currently experiencing rapid widespread warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme events [@zhang_increased_2023]. As we approach the deadline for the Kunming-Montreal Global Biodiversity Framework (i.e., the "30 by 30" initiative), a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], Canadian decision makers must understand which areas will be of greatest value for maximizing and protecting biodiversity not only for current conditions but future decades, too. Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term. Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022;  @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].

@@

We used the the `climatenaR` package [v. 1.0, @burnett_climatenar_2023] for `R` and ClimateNA [v. 7.4.2$;$ @wang_locally_2016; @mahony_global_2022] to obtain projected average monthly temperatures from 2020 to 2100 at a 0.08$^\circ$ spatial resolution. Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the historical data for western Canada from 1998 to 2023 (extremes included) to estimate the monthly variation in temperature, which we defined as $\sigma^2_T$, the variance at location $\langle x, y \rangle$ in month $m$ for each year (e.g., the variance at coordinates $\langle -119.40, 49.94 \rangle$ in April 2005). We then modeled the estimated variance using a Generalized Additive Model for Location and Scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017] using the `mgcv` package for `R` [v. 1.9-1, @wood_generalized_2017]. The linear predictor for the location (i.e., the mean) included smooth terms of: each year's estimated within-pixel monthly mean temperature ($\hat \mu_T$), month ($m$), a two-dimensional smooth of spatial coordinates $\langle x, y \rangle$, and a tensor product interaction term of month and space to allow for seasonal terms to vary smoothly over space. The linear predictor for the scale term, which governs the mean-variance relationship [table 3.1 on p. 104 of @wood_generalized_2017], included smooth terms of the monthly mean temperature, month, and space. We did not include smooth terms of year to avoid unrealistic projections when extrapolating beyond past 2023. The model was thus

@@

\noindent Predicted changes in movement rates with future climate change varied across species in both magnitude and direction, but worse SSPs always corresponded to greater absolute changes (Fig. \@ref(fig:t-dist)). Additionally, species that were predicted to move less often did not necessarily have lower speeds when moving, and vice versa (Figs. B9 and B10). Absolute relative changes in distance travelled were small under the best-case SSP (0-4% change in 2100 relative to 2025). Under the worst-case SSP, absolute changes by 2100 (relative to 2025) ranged from ~2% (grizzly bears) to ~24% (cougars). Notably, while the models estimated that grizzly bears would move substantially less (if at all) in late fall and winter (Figs. \@ref(fig:dist), B4), the models did not explicitly account for changes in hibernation phenology. Consequently, the climate change projections do not explicitly account for changes in energetic needs or physiology. Projected changes in 2100 varied spatially due to spatial heterogeneity in climate change projections (Fig. \@ref(fig:s-dist)). Again, absolute changes were generally greatest under worse SSPs, but the direction of change at each location also varied across SSPs (most visible in cougars).

@@

<!-- - address any questions presented in the introduction -->

\noindent We have demonstrated that temperature is an important determinant of how and where large mammals move, including how they alter seasonal and daily cycles in movement behavior. However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are increasingly frequently exposed to extreme and novel conditions.

***Temperature affects many aspects of mammalian physiology and behaviour, including energetics, sleep, and movement behaviour. Climatic changes during the last two centuries have exposed many mammals to novel and increasingly extreme environmental conditions that have led to visible changes that impact not only individuals but also ecological and human communities. However, quantifying the effects of climate change is often complex, especially as conditions become increasingly different and extreme events become increasingly common. Accurately quantifying the nonlinear effects of temperature on how and where mammals move requires careful data collection and model design. We have presented a multi-species analysis of the effects of temperature on mammalian movement rates and habitat selection. Understanding of how temperature affects each species in a community will allow us to begin making community-level inferences of how temperature will affect intra-specific interactions from a movement behaviour perspective [@nathan_movement_2008; @martinez-garcia_how_2020; @nathan_big-data_2022]. While adopting such a perspective is surely not simple, it would result in more complete estimates of the effects of climate change on community ecology, particularly when species are exposed to rapid change in many important variables. Future work should explore the effects of temperature on movement behaviour while accounting for finer-scale and species specific variables that were not accounted for in this study. Examples of these include temporally dynamic measures of forest type and age, canopy density, competitive pressures, forage availability, and predator avoidance, or environmental stochasticity. However, many of these data not available, so we also suggest that more work be done on quantifying such measures using spatiotemporally dynamic models and sufficiently fine resolutions.***

Predicting mammals' responses at the data-scarce fringes of the surface plots in Figs. \@ref(fig:dist) and \@ref(fig:hrsfs) comes with substantial uncertainty, particularly given that the responses are nonlinear. At warm temperatures, mammals overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures. As extreme temperature events become more frequent and prolonged throughout the year [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours that may alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023; @logares_black_2012; @anderson_black-swan_2017]. For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affect mammals' sleep quality and likelihood to enter torpor [@mortlock_sleep_2024; @fjelldal_nightly_2021].

Changes in mammal movement behaviour may also alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024]. At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023], increase their chances of finding food or being preyed on [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021]. These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011].

***Protecting a third of the world's ecosystems in accordance with the Kunming-Montreal Global Biodiversity Framework [@convention_on_biological_diversity_154_2022] appears less achievable as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023]. Achieving the "30 by 30" goal will require international collaboration [@huang_achieving_2021] as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023]. Understanding the consequences of climate change on mammals' movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018].***

\noindent The heterogeneity mammals' responses to temperature suggests that ecological communities will respond to change in complex and interconnected ways. Although our models do not account for explicit physiological or phenological changes, the surfaces in Fig. \@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns (most visible in cougars and grizzly bears). For example, when temperatures were above 0$^\circ$C, cougars moved most at night, but when temperatures were below 0$^\circ$C they tended to move more throughout the day. Throughout the year, they adapted their tolerance to temperature and moved less when it was relatively hot (for a given time of year), especially in spring and summer. The strong reduction in the mid-day movement rates of wolves, elk, cougars, and boreal caribou when summer temperatures were above 20$^\circ$C suggests that the increasingly common and intense heat waves across western Canada will have community-wide impacts on movement rates, encounter rates, and potentially community structure [@martinez-garcia_how_2020]. More work is necessary on quantifying interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024]. The increase in intensity and frequency of extreme heat events [@bunnell_global_2011; @yao_emergence_2022] will likely also impact the occurrence and timing of hibernation [@wells_life_2022] and migration or seasonal range expansions [@carbeck_adaptation_2022; @morley_spatiotemporal_2021; @malpeli_ungulate_2022]. However, not all species may be able to adapt at the current rate of climate change [@williams_range_2018; @hetem_responses_2014], especially since landscape changes will likely depend not only on the direct effect of temperature but also on cascading changes in the availability and unpredictability of resources [@mclellan_effect_2015; @pigeon_drivers_2016; @mezzini_how_2025].

***Overall, estimated selection strength was weakest for forest cover and strongest for elevation. Elevation is the only raster that doesn't change noticeably over time, so time-varying rasters of the other two would have likely given stronger (and better) estimates.***

As with the effects of temperature on movement rates, the hererogeneity of temperature effects on species' habitat selection indicates that efficient and proactive habitat conservation will require taxonomic nuance. In particular, our HRSFs did not account for any explicit physiological responses to temperature that may impact movement, such as changes in the phenology of plants, fire, ice, or mammalian physiology (e.g., moulting, fat storages) or behaviour (e.g., migration, food caching), other than as direct responses to temperature. Consequently, the behavioural changes we present should be interpreted carefully. While some seasonal cycles do not depend directly on temperature and may thus not be impacted directly by climate change [such as those driven by photoperiod, see @liu_effects_2022], the predictions we present also rely on the assumption that mammals will not adapt these seasonal cycles to temperature-dependent cues, such as weather [@xu_frost_2019] or forage availability [@middleton_green-wave_2018], which may not be correct. Additionally, it is important to stress the distinction between habitat selection and quality. While individuals tend to select for high-quality habitats [@kacelnik_ideal_1992; @wright_survival_2024], high selection strength can also be a consequence of an ecological trap [@zuniga-palacios_what_2021; @swearer_overview_2021].

* geographic space *sensu* @matthiopoulos_defining_2022
* environmental space *sensu* @matthiopoulos_defining_2022
* G space is the realization of the theoretical E space

## Predicted changes in movement during the 21^st^ century

Achieving the goals laid out by the "30 by 30" conservation initiative [@convention_on_biological_diversity_154_2022] will require careful planning but often also rapid action. As of December 2024, only 13.8% of terrestrial Canada was within protected areas or other conserved areas [@environment_and_climate_change_canada_canadian_2025]. Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [@falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] and as we actively prepare for future change [@desjardins_looking_2020; @hessami_indigenizing_2021].

***...*** spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas [@jeltsch_need_2025]. Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024; @silva_too_2025]. Additionally, accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019]. This complexity results in the great variation among the responses of different species (and individuals), especially as populations undergo a rapid cascade of change [@botero_evolutionary_2015] and mammals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022]. As selection strength for current ranges changes (and likely decreases), some individuals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased risk of extirpation or extiction [@logares_black_2012; @anderson_black-swan_2017; @duncan_predicting_2012].

## Species-specific considerations

* Cougars: attraction for disconnected patches of high elevation could lead to fragmentation or traps

* boreal caribou: Increasing temperatures increased RSS for both elevations near ~500 m as well as > 1,000 m, but such a selection is only possible if: (1) such habitats exist in the individuals' range, and (2) crossing the 500 m to > 1,000 m regions is safe and attractive

* sm caribou: HRSFs do not account for migratory behaviour due to reproduction or predation risk

## Study limitation and considerations for future studies

Our work highlights three central considerations for future research: (1) telemetry sampling rates should be designed primarily in relation to the movement timescales of the species of interest [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @pease_ecological_2024], (2) the number of individuals tracked and the duration of each telemetry should depend not just on population size but also the rate of environmental change, and (3) predicting changes in movement behaviour becomes highly complex when responses are nonlinear, especially when changes depend on many interacting factors [@polazzo_measuring_2024] and one is interested in predicting responses in extreme conditions for which data are scarce [@steixner-kumar_strategies_2020].

>>>>

Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010]. While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more severe physiological damage in a shorter span of time [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024]. Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, especially with increasing body size [@dyer_travel_2023; @verzuh_behavioural_2023]. Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature genralists') can adapt more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024]. Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond within a lifetime or a few generations [@levins_evolution_1974; @gienapp_climate_2008; @fuller_towards_2016; @xu_plasticity_2021].

@@

In the last century, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ipcc_climate_2023]. As heat stress intensifies over the next 100 years, mammals will continue to suffer impacts on their fitness, movement rates, and habitat selection [@deb_modelling_2020; @woo-durand_increasing_2020], which will be exacerbated by more frequent and intense extreme events [@bunnell_global_2011] and anthropogenic habitat loss, modification, and fragmentation [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024]. It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018]. Preparing for and responding to future changes require a comprehensive understanding of how temperature will affect how and where mammals move, as well as the implications such changes will have on species' adaptability and overall community structure [@cunningham_opportunity_2021]. Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], thermoregulation [@mota-rojas_physiological_2021], and movement behaviour [@mccain_assessing_2019; @melin_effect_2023; @cunningham_opportunity_2021]. However, there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behaviour and quantitative predictions of changes in movement behaviour with climate change in the current century [@ipcc_climate_2023].

@@

Western Canada is currently experiencing rapid widespread warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme events [@zhang_increased_2023]. As we approach the deadline for the Kunming-Montreal Global Biodiversity Framework (i.e., the "30 by 30" initiative), a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], Canadian decision makers must understand which areas will be of greatest value for maximizing and protecting biodiversity not only for current conditions but future decades, too. Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term. Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022;  @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].

@@

We used the the `climatenaR` package [v. 1.0, @burnett_climatenar_2023] for `R` and ClimateNA [v. 7.4.2$;$ @wang_locally_2016; @mahony_global_2022] to obtain projected average monthly temperatures from 2020 to 2100 at a 0.08$^\circ$ spatial resolution. Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the historical data for western Canada from 1998 to 2023 (extremes included) to estimate the monthly variation in temperature, which we defined as $\sigma^2_T$, the variance at location $\langle x, y \rangle$ in month $m$ for each year (e.g., the variance at coordinates $\langle -119.40, 49.94 \rangle$ in April 2005). We then modeled the estimated variance using a Generalized Additive Model (GAM) for Location and Scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017] using the `mgcv` package for `R` [v. 1.9-1, @wood_generalized_2017]. The linear predictor for the location (i.e., the mean) included smooth terms of: each year's estimated within-pixel monthly mean temperature ($\hat \mu_T$), month ($m$), a two-dimensional smooth of spatial coordinates $\langle x, y \rangle$, and a tensor product interaction term of month and space to allow for seasonal terms to vary smoothly over space. The linear predictor for the scale term, which governs the mean-variance relationship [table 3.1 on p. 104 of @wood_generalized_2017], included smooth terms of the monthly mean temperature, month, and space. We did not include smooth terms of year to avoid unrealistic projections when extrapolating beyond past 2023. The model was thus

@@

\noindent Predicted changes in movement rates with future climate change varied across species in both magnitude and direction, but worse SSPs always corresponded to greater absolute changes (Fig. \@ref(fig:t-dist)). Additionally, species that were predicted to move less often did not necessarily have lower speeds when moving, and vice versa (Figs. B9 and B10). Absolute relative changes in average yearly distance travelled were small under the best-case SSP (0-4% change in 2100 relative to 2025). Under the worst-case SSP, absolute changes by 2100 (relative to 2025) ranged from ~2% (grizzly bears) to ~24% (cougars). Notably, while the models estimated that grizzly bears would move substantially less (if at all) in late fall and winter (Figs. \@ref(fig:dist), B4), the models did not explicitly account for changes in hibernation phenology. Consequently, the climate change projections do not explicitly account for changes in energetic needs or physiology. Projected changes in 2100 varied spatially due to spatial heterogeneity in climate change projections (Fig. \@ref(fig:s-dist)). Again, absolute changes were generally greatest under worse SSPs, but the direction of change at each location also varied across SSPs (most visible in cougars).

@@

<!-- to do: address any questions presented in the introduction -->

\noindent We have demonstrated that temperature is an important determinant of how and where large mammals move, and that mammals alter their seasonal and daily cycles in movement behavior in response to changes in temperature. Predicting mammals' responses to climate change remains a challenge, as habitats warm rapidly and mammals are exposed to increasingly novel and extreme conditions [@ipcc_climate_2023; @diffenbaugh_changes_2013; @botero_evolutionary_2015]. We leveraged the flexibility and interptetability of GAMs to estimate the (nonlinear) responses  of terrestrial mammals to temperature. In particular, tensor product interaction terms were crucial for estimating smooth, nonlinear changes in daily and seasonal trends with temperature. By allowing interaction terms to be smooth bivariate functions, we were able to quantify changes in movement behavior without the need to discretize time into distinct windows (e.g., day/night and seasons). We were thus able to present changes using relatively simple and interpretable surface plots (Figs. \@ref(fig:dist) and \@ref(fig:hrsfs)) that conveyed detailed and nuanced estimates of change. This framework was essential for estimating how mammals behave in extreme conditions for which data are scarce.

As temperatures warm, mammals' risk of heperthermia increases nonlinearly [@sherwood_adaptability_2010], especially for large-bodied species [@alston_temperature_2020; @dyer_travel_2023; @verzuh_behavioural_2023]. The recent increase in frequency and duration of extreme-temperature events [@yao_emergence_2022; @ipcc_climate_2023; @zhang_increased_2023; @bunnell_global_2011; @kienzle_has_2018] have forced many mammals towards increasingly atypical behaviours with long-term consequences, ranging from changes in the behavior and fitness of individuals [@brivio_seeking_2024; @cunningham_opportunity_2021] to mass die-offs [@ameca_y_juarez_natural_2012]. A species' ability to respond to changing temperature and phenology depends mainly on the its exposure, sensitivity, and adaptability to changes, and growing evidence suggests that many mammals in western Canada will struggle to respond due limitations in all three factors [@ameca_y_juarez_natural_2012; @diffenbaugh_changes_2013]. Firstly, exposure to warming temperatures is widespread throughout the area, and warming is expected to accelerate over the coming decades [@ipcc_climate_2023; @diffenbaugh_changes_2013; @bunnell_global_2011; @kienzle_has_2018; @eum_potential_2023]. Secondly, large-bodied mammals are particularly sensitive to high heat [@dyer_travel_2023], but small-bodied mammals are also affected [@ratnayake_forecasting_2019; @krebs_impact_2019]. Finally, species' adaptability is tested by the speed of climate change [@diffenbaugh_changes_2013] and the rigidity of many speices' photoperiod-dependent phenology [@walker_global_2019].

<!-- complex response from each species -->

Our results suggest that climate change will cause mammals to alter their seasonal movement behavior, but not all species will respond similarly. For example, hotter temperatures caused boreal caribou to travel longer distances in fall, winter, and spring but less in summer, mirroring the findings of @stien_congruent_2012, @leclerc_determinants_2021, and @lessard_local_2025. The increase in movement rates during the colder seasons are likely partly due to shallower snow depth that results in greater mobility [@pedersen_quantifying_2021], but rising temperatures during snowy seasons also increase the chances of rain-on-snow events, which limit forage availability and increase the ungulates' need to search for food [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025]. In contrast, the reduced movement rates during warmer summers are likely mainly due to caribou searching for shelter from the heat due to ungulates' elevated risk of hyperthermia [@alston_temperature_2020; @verzuh_behavioural_2023]. The effects of extreme temperatures are exacerbated by phenological mismatches with seasonal photoperiod [@walker_global_2019], including the timing of molting and reproduction. Earlier growth seasons in boreal and arctic regions have resulted in lower calf birth and survival [@post_climate_2008], while the lengthening of the growth season has allowed moose (*Alces alces*) and deer (*Odocoileus* spp.) to encroach on boreal caribou habitat, along with increased density of coyotes (*Canis latrans*), cougars, and wolves [@barber_potential_2018; @demars_incorporating_2023]. Thus, rising temperatures affect more than each individual species' movement behavior; they have complex and interconnected effects on an entire community and its trophic structure.

<!-- complex response from community -->

Our results suggest that rising temperatures will have similar effects on boreal wolves. Warmer temperatures caused wolves to travel further throughout the year, but, similarly to boreal caribou, they moved substantially less at temperatures above 25$^\circ$C in summer. At temperatures above 20$^\circ$, both boreal caribou and wolves tended to move less during the day and more between 17:00 and 24:00. This synchroniciy in movement patterns is likely to increase encounters between the two species, particularly when coupled with increasing human disturbances, since wolves leverage seismic lines and other anthropogenic linear features (e.g., roads) to reduce movement costs while increasing the chances of encountering prey [@whittington_caribou_2011; @dickie_faster_2017; @dickie_resource_2022]. Additionally, caribou that attempt to reduce predation risk from wolves by avoiding wolf habitat may risk increasing predation pressure from bears [@leblond_caribou_2016] and other encroaching predators [@demars_incorporating_2023]. As temperatures rise and allow new competitors and predators to encroach on boreal caribou habitat [@barber_potential_2018; @demars_incorporating_2023; @labadie_global_2023], climate change will have complex conseuqences not only for individual species but for entire communities and their trophic interactions.

<!-- focus on habitat selection using changes in elevational preference of elk, goats, and cougars as the main example -->

Considerations about changes in trophic interactions leads to an important caveat about the results we present. The estimated effects of temperature on movement behavior cannot be fully attributed to physiological responses to changes in temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature. For example, mountain goats' reduced movement rates at temperatures near 0$^\circ$C in winter may be due to increased movement difficulty, since melting snow and rain-on-snow events affect the energetic costs of movement [@white_population_2025] and encounter rates with predators [@sullender_defining_2023]. However, milder temperatures can also increase plant growth, which, in turn, allows goats to spend less time searching for forage [@white_mountain_2025]. Therefore, the figure \@ref(fig:dist) illustrates an estimate of the total effect of temperature on each species movement rates, but it does not explicitly account for energetic costs. Similarly, species' changes in habitat selection do not explicitly account for any changes in physiological or energetic costs. Both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not explicitly account for differences in forage availability at different altitudes.

Forage quality and availability is often a limiting factor for mammals' ability to adapt to warming temperatures by moving to higher altitudes. Plants are slower to arrive in and establish in new habitats than mammals, and the advance of herbivores can further limit plants' ability to establish [@speed_elevational_2012; @diffenbaugh_changes_2013]. Consequently, mammals that move to higher elevations may be forced to spend more time foraging [@mezzini_how_2025], which will increase their energetic expenditure as well as their chances of encountering predators (including cougars and grizzly bears) or competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023]. Additionally, plants at high elevations may be have weaker defenses against herbivory and be more susceptible to overgrazing [@callis-duehl_community-level_2017]. The scarcity and susceptibility of forage may thus limit herbivores' ability to shift elevationally, particularly if predation rates remain relatively unchanged or increase. Furthermore, the physical upper limit to elevation can cause individuals to become trapped . If local peaks become unsuitable habitat, and moving to nearby higher peaks requires traveling across low-quality and high-risk habitat (Figure \@ref(fig:s-rss-cc)), habitat fragmentation may become too excessive to move across patches or alternative ranges [@white_mountain_2025]. Translocating animals to locations with higher peaks may help avoid such a scenario, but the success of relocations can be highly species-dependent, and attempts are not always successful [@ranc_memory_2022; @white_mountain_2025].

Changes in mammals' seasonal movement behavior are also likely to affect the timing and frequency of human-wildlife interactions. Since heat affects mammals' sleep quality and likelihood to enter torpor [@mortlock_sleep_2024; @fjelldal_nightly_2021; @wells_life_2022], future decades may see increased levels of incidents with bears that struggle to enter or remain in hibernation [@johnson_human_2018]. The risk is further increased when bears' are unable to meet their increased energetic requirements through foraging or hunting. Projected increases in park attendance due to rising temperatures and human population density in Western Canada are expected to further increase frequency of human-wildlife interactions [@weststrate_how_2024].

## Consequences for conservation

Species' adaptability to rising temperatures is challenged by the aforementioned symultaneous, multifarious, and interconnected stressors caused by climate change, since responding to many concurrent changes at once is particularly difficult [@diffenbaugh_changes_2013; @polazzo_measuring_2024]. Locations may thus not be able to support species and population sizes they supported historically, and species may thus not select for historical geograpical ranges in the future. It is therefore crucial for conservationists and other decision-makers to distinguish between a species' ideal environmental space [or E space, *sensu* @matthiopoulos_defining_2022], and its best available geographic space [or G space, *sensu* @matthiopoulos_defining_2022]. Accelerating multufarious change in species' movement behavior and energetic costs require conservation measures to be proactive and anticipate future changes, even when future conditions are substantially different from historical conditions and historical data may not be available. To this end, the framework presented by @polazzo_measuring_2024 for estimating species' responses to multifarious change becomes essential, particularly when accounting for energetic and fitness costs, as argued by @cunningham_opportunity_2021.

As of December 2024, only 13.8% of terrestrial Canada was within protected areas or other conserved areas [@environment_and_climate_change_canada_canadian_2025], so rapid action is necessary to protect and additional 16.2% of Canada's land by 2030 in accordance with Kunming-Montreal Global Biodiversity Framework [i.e., the "30 by 30 initiative"$;$ @convention_on_biological_diversity_154_2022]. However, the choice of what habitats to protect will require careful consideration, since protected areas should support species not only in present conditions but also as species' needs and selection change over the coming decades [@simmons_tough_2021]. As anthropogenic activity increasingly alters habitats and community structures [@sih_evolution_2011; @tucker_moving_2018; @rosenthal_impact_2022; @rice_conspicuous_2022], protecting biodiversity and the habitats it depends on will require widespread and collaboration with local stakeholders, especially Indigenous Peoples, to ensure that all decisions are culturally sound and respect Treaties [@turner_its_2009; @desjardins_looking_2020; @wong_towards_2020; @hessami_indigenizing_2021; @falardeau_biophysical_2022; @lamb_braiding_2023]. Understanding the consequences of climate change in mammalian movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018]. However, a better understanding of such consequences will require a more holistic approach that includes carefully data collection, model design, and interpretation and application of results.

## Study limitations and considerations for future work

***HERE***

Estimated selection strength for forest cover was weak and did not change strongly with temperature. Time-varying rasters of forest cover would have likely given stronger (and better) estimates.

* variation among individuals: @jeltsch_need_2025. Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024; @silva_too_2025].

- we see clear signals despite the data issues and limitations, but more work needs to be done to disentangle important drivers

More work is necessary on quantifying interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024]. The increase in intensity and frequency of extreme heat events [@bunnell_global_2011; @yao_emergence_2022] will likely also impact the occurrence and timing of hibernation [@wells_life_2022] and migration or seasonal range expansions [@carbeck_adaptation_2022; @morley_spatiotemporal_2021; @malpeli_ungulate_2022]. However, not all species may be able to adapt at the current rate of climate change [@williams_range_2018; @hetem_responses_2014], especially since landscape changes will likely depend not only on the direct effect of temperature but also on cascading changes in the availability and unpredictability of resources [@mclellan_effect_2015; @pigeon_drivers_2016; @mezzini_how_2025].
